Finiteness in Cubical Type Theory

### 2020-01-21 paper.tex
more

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
While proof search based on finiteness is not new, implementing it here does
give us an opportunity to demonstrate some actual \emph{computation} in a
univalent setting.
Many of the finiteness predicates are built using univalence, so for the proof
search functionality to work at all we require computational content in the
univalence axiom.
Furthermore, the extensionality afforded to us by HoTT means that we can have
\(\Pi\) types in the domain of the search, meaning we can synthesise full
functions from specifications alone.
\subsection{Related Work}
\subsection{Notation and Background}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into type.
  \item[Universes]
    Implicitly, types will exist at some point in a universe hierarchy,
    beginning with \(\mathbf{Type}\) (\(\mathbf{Type} : \mathbf{Type}_1\),
    \(\mathbf{Type}_1 : \mathbf{Type}_2\), and so on).
    In our formalisation, every proof is done in the most universe polymorphic
    way possible: the proof that the cardinally finite sets form a
    \(\Pi\)-pretopos, for instance, is defined over any universe level.
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] Disjoint union can be defined in terms of \(\Sigma\):
    \begin{equation}
      A \uplus B \coloneqq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then } A \text{ else } B
    \end{equation}
    However, we prefer to use it as an inductively defined type (though the two
    are equivalent).
    The two constructors are:
    \begin{equation}
      \begin{aligned}
        \text{inl} &: A \rightarrow A \uplus B \\
        \text{inr} &: B \rightarrow A \uplus B
      \end{aligned}
    \end{equation}
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage.
\end{description}

@@

\todo[inline]{Basic MLTT Notation: \(\Pi\), \(\Sigma\), etc}
\todo[inline]{Paths, Univalence, isomorphisms}
\todo[inline]{HITs}
\todo[inline]{Surjections}
\todo[inline]{Containers}
\todo[inline]{Fin}
\todo[inline]{Biconditional, Isomorphism, Equivalence, Equals}
\section{Finiteness Predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
The reason we explore predicates other than our focus (cardinal finiteness) is
that we can often prove things like closure much more readily on the simpler
predicates.
The relations (which we will prove in the next section) then allow us to
transfer those proofs onto Kuratowski finiteness.
\subsection{Split Enumerable}
\begin{romdefinition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{romdefinition}

>>>>

All of our work is formalised in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
We will make mention of the few occasions where the formalisation of some proof
is of interest, but the main place where we will discuss the code is in the
final section, where we implement a library for proof search, based on
omniscience and exhaustibility.
While proof search based on finiteness is not new, implementing it here does
give us an opportunity to demonstrate some actual \emph{computation} in a
univalent setting.
Many of the finiteness predicates are built using univalence, so for the proof
search functionality to work at all we require computational content in the
univalence axiom.
Furthermore, the extensionality afforded to us by HoTT means that we can have
\(\Pi\) types in the domain of the search, meaning we can synthesise full
functions from specifications alone.
\subsection{Related Work}
\subsection{Notation and Background}
\subsubsection{Notation}
We work in Cubical Type Theory \cite{cohenCubicalTypeTheory2016}.
For the various type formers we use the following notation:
\begin{description}
  \item[Type] We use \(\mathbf{Type}\) to denote the universe of (small) types.
    ``Type families'' are functions into type.
  \item[Universes]
    Implicitly, types will exist at some point in a universe hierarchy,
    beginning with \(\mathbf{Type}\) (\(\mathbf{Type} : \mathbf{Type}_1\),
    \(\mathbf{Type}_1 : \mathbf{Type}_2\), and so on).
    In our formalisation, every proof is done in the most universe polymorphic
    way possible: the proof that the cardinally finite sets form a
    \(\Pi\)-pretopos, for instance, is defined over any universe level.
  \item[0 , 1 , 2] We call the \(\mathbf{0}\), \(\mathbf{1}\), and
    \(\mathbf{2}\) types \(\bot\), \(\top\), and \(\mathbf{Bool}\) respectively.
    The single inhabitant of \(\top\) is tt, and the two inhabitants of
    \(\mathbf{Bool}\) are false and true.
  \item[Dependent Sum and Product] We use \(\Sigma\) and \(\Pi\) for the
    dependent sum and product, respectively.
    The two projections from \(\Sigma\) are called fst and snd.
    In the non-dependent case, \(\Sigma\) can be written as \(\times\), and
    \(\Pi\) as \(\rightarrow\).
  \item[Disjoint Union] Disjoint union can be defined in terms of \(\Sigma\):
    \begin{equation}
      A \uplus B \coloneqq \Sigma(x : \mathbf{Bool}) , \text{if } x \text{ then } A \text{ else } B
    \end{equation}
    However, we prefer to use it as an inductively defined type (though the two
    are equivalent).
    The two constructors are:
    \begin{equation}
      \begin{aligned}
        \text{inl} &: A \rightarrow A \uplus B \\
        \text{inr} &: B \rightarrow A \uplus B
      \end{aligned}
    \end{equation}
  \item[Equalities, equivalences, and paths] We use the symbol \(\coloneqq\)
    for definitions.
    \(\simeq\) will be used for equivalences, and \(\equiv\) for equalities.
    Of course, we know that \((A \simeq B) \simeq (A \equiv B)\) by univalence,
    so the distinction isn't terribly important in usage.
\end{description}
\subsubsection{Cubical Type Theory}
Cubical Type Theory \cite{cohenCubicalTypeTheory2016} is a constructive type
theory with an implementation in Cubical Agda
\cite{vezzosiCubicalAgdaDependently2019}.
It allows us to do much of the same theory as in HoTT, but crucially the
univalence ``axiom'' is a \emph{theorem}, rather than an axiom.
This allows us to actually compute with univalent proofs, a capability missing
from HoTT.
\begin{romdefinition}[Path Types]
  The equality type (which we denote with \(\equiv\)) in CuTT is the type of
  Paths\footnotemark.
  The internal structure of paths is largely irrelevant to us here, as we will
  generally treat \(\equiv\) as a black-box equivalence relation with
  substitution and congruence.
\end{romdefinition}

\footnotetext{
  Actually, CuTT does have an identity type with similar semantics to the
  identity type in MLTT.
  We do not use this type anywhere in our work, however, so we will not consider
  it here.
}
\begin{romdefinition}[Homotopy Levels]
  Types in HoTT and CuTT are not necessarily sets, as they are in MLTT.
  This means that equalities are not necessarily unique.
  In fact, we have an entire hierarchy of homotopy levels, of which sets are
  level 2.
  \begin{align}
    \text{isContr}(A)    &\coloneqq \Sigma(x : A) , \Pi(y : A) , (x \equiv y) \\
    \text{isProp}(A)     &\coloneqq \Pi(x, y : A) , (x \equiv y) \\
    \text{isSet}(A)      &\coloneqq \Pi(x, y : A) , \text{isProp}(x \equiv y) \\
    \text{isGroupoid}(A) &\coloneqq \Pi(x, y : A) , \text{isSet}(x \equiv y)
  \end{align}
  We can define the above types inductively like so:
  \begin{align}
    \text{isOfHLevel}(0 , A) &\coloneqq \text{isContr}(A) \\
    \text{isOfHLevel}(n+1 , A) &\coloneqq \Pi(x , y : A) , \text{isOfHLevel}(n , x \equiv y)
  \end{align}

\end{romdefinition}

\begin{romdefinition}[Isomorphism]
  We say two types \(A\) and \(B\) are isomorphic, denoted with \(A \iff B\), if
  there is a pair of functions \(f : A \rightarrow B\), \(g : B \rightarrow A\)
  such that
  \begin{equation}
    \begin{aligned}
      \Pi(x : A) , (g(f(x)) \equiv x) \\
      \Pi(x : B) , (f(g(x)) \equiv x)
    \end{aligned}
  \end{equation}
  While we can derive an equivalence from an isomorphism:
  \begin{equation}
    \text{isoToEquiv} : (A \iff B) \rightarrow (A \simeq B)
  \end{equation}
  It is not the case that an isomorphism is the \emph{same} as an equivalence in
  all cases.
  In other words, we do not have:
  \begin{equation}
    (A \iff B) \simeq (A \simeq B)
  \end{equation}
  However, if one of \(A\) or \(B\) is a set, then the above holds.
\end{romdefinition}

@@

\todo[inline]{HITs}
\todo[inline]{Surjections}
\todo[inline]{Containers}
\todo[inline]{Fin}
\todo[inline]{Biconditional, Isomorphism, Equivalence, Equals}
\section{Finiteness Predicates}
In this section, we will define and briefly describe each of the five predicates
in Figure~\ref{finite-classification}.
The reason we explore predicates other than our focus (cardinal finiteness) is
that we can often prove things like closure much more readily on the simpler
predicates.
The relations (which we will prove in the next section) then allow us to
transfer those proofs onto Kuratowski finiteness.
\subsection{Split Enumerable}
\begin{romdefinition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in \mathit{xs}
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
  An equivalent version of this predicate was called \verb+Listable+ in
  \cite{firsovDependentlyTypedProgramming2015}.
\end{romdefinition}
